Optimal updating factor in adaptive flat-distribution simulations

### 2016-04-22 doc/invt.tex
add new figure errinvt.gp

\title{Optimal updating factor in adaptive flat-distribution-sampling simulations}

@@

\subsubsection{\label{sec:nnscheme}
Nearest-neighbor updating scheme}

@@

We first tested the theory of computing errors
on the simple inverse-time schedule,
%
\begin{equation}
\alpha(t) = \frac{1}{\lambda \, (t + t_0) },
\label{eq:alpha_invtlambda}
\end{equation}
%
where $\lambda$ is a free parameter,
and $t_0 = 2/(\lambda \, \alpha_0)$.
%
The error of this schedule
can be computed analytically
(cf. Appendix \ref{sec:invt_schedule}),
and tested against simulation.

%\subsubsection{Single-bin (WL) updating scheme}

For the single-bin updating scheme, we
%we chose $\alpha_0 = 10^{-4}$ and
ran the simulation for $T = 10^8$ steps.
%
As shown in Fig. \ref{fig:errsbin},
the numerical results of the terminal error agreed well with
the analytical prediction.
%
For both global and local sampling processes,
the optimal $\lambda$ occurred around $1.0$,
verifying the optimality of Eq. \eqref{eq:alpha_invt1}.
%
%Recall that $1/\lambda$ in Eq. \eqref{eq:alpha_invtlambda}
%represents the relative updating magnitude.
%%
%Figure \ref{fig:errsbin} also shows
%an asymmetry of the error curve.
%A smaller than optimal value of
%the updating magnitude, $1/\lambda$,
%tends to increase the error more rapidly
%than a larger than optimal value,
%suggesting that the over-updating
%is generally preferable to the under-updating
%when the optimal value is unknown.
%
The two error functions
differed only by a multiplicative constant,
also in agreement with the theory
(cf. Appendix \ref{sec:invt_singlebin}).
%

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/errsbin.pdf}
  }
  \caption{
    \label{fig:errsbin}
    Error, $E$, versus the
    relative updating magnitude,
    %proportionality constant,
    $1/\lambda$,
    in Eq. \eqref{eq:alpha_invtlambda}
    for the single-bin (WL) scheme.
    %
    The results were averaged over $1000$ independent runs.
    %
    \note{The figure was produced by \texttt{doc/fig/errsbin.gp}.
      The data were produced by the script \texttt{data/invt2run.py},
      and saved under \texttt{data/singlebin}
      (see the \texttt{Makefile} there).
    }%
  }
\end{center}
\end{figure}

%\subsubsection{Nearest-neighbor updating scheme}

As an example of multiple-bin schemes,
we tested the nearest-neighbor updating scheme
described in Sec. \ref{sec:nnscheme},
%
with
$\mu_1 = 0.24$ to make the smallest eigenvalue
close to zero.
%
Again, simulation and theory agreed well,
as shown in Fig. \ref{fig:errnnbr}.
%
The optimal value of $1/\lambda$,
increased with the simulation length, $T$,
in the global sampling case
(cf. Appendix \ref{sec:invt_deplength}).
%
However, the shift was suppressed
for the local sampling case,
with the optimal $\lambda$ remained around $1.0$
[cf. Eq. \eqref{eq:lambda_nn_onestep}].

%To understand this dependence,
%we observe that for shorter simulations,
%long-wavelength modes
%(with larger $\lambda_k$ and smaller $k$),
%dominate the error function in Eq. \eqref{eq:error_invt}
%because of the prefactor $\lambda_k \, \Gamma_k$,
%which is inherited from
%the equilibrium error under a fixed updating magnitude,
%Eq. \eqref{eq:error_eql}.
%%
%For longer simulations,
%short-wavelength modes
%(with smaller $\lambda_k$ and larger $k$)
%become more important,
%for $\lambda_k$ also gives the decay rate
%of the $k$th mode.
%%and errors in smaller $\lambda_k$ modes
%%are harder to damp out.
%%
%This drives the optimal $\lambda$ to shift toward
%a smaller $\lambda_k$ value
%as the simulation lengthens.
%%
%Note also that while increasing $\lambda$
%also increases the error of the long-wavelength modes,
%this rate of increase is much smaller,
%as shown in the $\nu_k > 1$ branch
%in Fig. \ref{fig:err_component}.
%%
%The shift of $\lambda$, however, is less noticeable
%under the local sampling scheme,
%because longer-wavelength modes more heavily
%weighted by much larger values of $\Gamma_k$.

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=0.95\linewidth]{fig/errnnbr.pdf}
  }
  \caption{
    \label{fig:errnnbr}
    Normalized error, $(T + t_0) \, E$,
    versus the %proportionality constant,
    relative updating magnitude,
    $1/\lambda$,
    in Eq. \eqref{eq:alpha_invtlambda}
    for the nearest-neighbor updating scheme with $\mu_1 = 0.24$.
    %
    The results were averaged over $1000$ independent runs.
    %
    \note{This figure was produced by \texttt{doc/fig/errnnbr.gp}.
      The data were produced by \texttt{data/invt2run.py},
      saved under \texttt{data/nb0.24}
      (see the \texttt{Makefile} there).
    }%
  }
\end{center}
\end{figure}

%\subsection{Gaussian (metadynamics) updating scheme}
%
%
%The error of Gaussian updating scheme is similar to
%that of the nearest neighbor scheme,
%as shown in Fig \ref{fig:err_sig5}.
%%
%The dependency of the simulation length is more prominent.
%%
%Even for local sampling,
%there is now a visible shift of the optimal $\lambda$.
%
%
%\begin{figure}[h]
%\begin{center}
%  \makebox[\linewidth][c]{
%    \includegraphics[angle=0, width=0.95\linewidth]{fig/errsig5.pdf}
%  }
%  \caption{
%    \label{fig:err_sig5}
%    Normalized error, $(T + t_0) \, E$,
%    versus the %proportionality constant,
%    relative updating magnitude,
%    $1/\lambda$,
%    in Eq. \eqref{eq:alpha_invtlambda}
%    for the Gaussian updating scheme with $\sigma = 5$ bins.
%    %
%    The results were averaged over $1000$ independent runs.
%  }
%\end{center}
%\end{figure}
%

@@

\subsection{\label{sec:results_cmpschemes}
Comparison of updating schemes}

@@

%A practical problem of using homogeneous updating scheme
%is the following.
%
A homogeneous updating scheme
specified by an arbitrary kernel
is not necessarily stable,
i.e., some eigenvalues may be negative.
%
%
%\note{
  %For example,
  %the nearest-neighbor updating scheme
  %is unstable if Eq. \eqref{eq:nn_stable} is violated.
  %
  Even
  the Gaussian updating scheme can be unstable
  if the updating kernel is truncated.
%}%
%
Here we show how to minimally modify
the updating kernel
to stabilize it.
%
\note{The relative code is \texttt{trimwindow()} in \texttt{invt.h}.
}

@@

For the single-bin scheme, we have
$\lambda_2 = \cdots = \lambda_n = 1$,
and
the errors
%from Eqs. \eqref{eq:error_res_invt}
%and \eqref{eq:error_asym_invt}
depend on $\Gamma_k$'s only via the sum
$\sum_{k = 2}^n \Gamma_k$,
which serves as a multiplicative constant.

@@

\subsection{\label{eq:invt_nn}
Nearest-neighbor updating scheme}

For the nearest-neighbor updating scheme
introduced in Sec. \ref{sec:nnscheme},
we can find an analytical expression
of the optimal value of $\lambda$
in the large-$n$ and asymptotic limit.
%
In this case,
if $2 \, \nu_k > 1$ for every $k$,
the asymptotic error is given by the integral
%$\sum_i \to \frac{2 \, n}{\pi} \int_0^{\pi/2} dp$, and
%
\begin{align}
  \Err_A
  =
  \frac{ 2 \, n  }
       { \pi \, T }
  \int_0^{\pi/2}
    \frac{ \Gamma(p) \, \left(1 - 4 \, \mu_1 \, \sin^2 p \right)^2    }
         {   \lambda \, \left(2 - 8 \, \mu_1 \, \sin^2 p - \lambda \right) }
  \, dp
  .
\notag
%\label{eq:error_nn_asym_int}
\end{align}

For perfect sampling,
$\Gamma(p) = 1$ from Eq. \eqref{eq:Gamma_perfect},
and
$$
\begin{aligned}
  \Err_A
  &=
  \frac{n}{4 \, T}
  \left(
    \frac{2 - 4 \, \mu_1 + \lambda}{ \lambda }
    +
    \frac{ \lambda }
    { \sqrt{ (2 - \lambda) (2 - 8 \, \mu_1 - \lambda) } }
  \right)
.
\end{aligned}
$$
%
Minimizing the function yields
%
\begin{equation}
  \lambda
  =
  \frac{ 1 - 4 \, \mu_1 }
       { 1 - 2 \, \mu_1 }
  ,
\notag
%\label{eq:lambda_nn_perfect}
\end{equation}
%
and the asymptotic error in the optimal case is given by
%
\begin{equation}
  \Err_A
  =
  \frac{n}{T}
  \left(
    1+ \frac{2 \, \mu_1^2}{1-4 \, \mu_1}
  \right)
  ,
\notag
%\label{eq:error_nn_perfect}
\end{equation}
%
%From Eq. \eqref{eq:error_nn}, it is clear that
which
increases with the magnitude of $\mu_1$
no matter its sign,
and the minimal value, obtained at $\mu_1 = 0$,
corresponds to the single-bin scheme.
%

For the one-step sampling process,
we find from Eq. \eqref{eq:Gamma_onestep}
that $\Gamma(p) = \cot^2 p$
is peaked around $p \approx 0$,
%
Then
$$
  \Err_A
  \propto
  \frac{   2 \, n }
       { \pi \, T }
  \frac{             1             }
       {  \lambda \, (2 - \lambda) }
  .
$$
%
The optimal $\lambda$ is, therefore,
%
\begin{equation}
\lambda \approx 1.
\label{eq:lambda_nn_onestep}
\end{equation}
%

However, Eq. \eqref{eq:lambda_nn_onestep}
is an approximation good only
for sufficiently short simulations,
since we have taken the limit of $n \to \infty$
for a fixed $T$.
%
For very long simulations,
the optimal $\lambda$
will ultimately drift to
the smallest eigenvalue,
$\min \lambda_k = 1 - 4 \, \mu_1$.

\subsection{\label{sec:invt_deplength}
Dependence on the simulation length}

For a finite-length simulation,
we generally expect the optimal $\lambda$
that minimizes the total error to
decrease with the simulation length $T$.
%
From Eq. \eqref{eq:alpha_invt}, we have
$$
  \lambda
  =
  \Delta \left( \frac{ 1 } { T \, \alpha } \right)
  \equiv
  \left.
  \frac { 1 } { T \, \alpha(t) }
  \right|_{t = 0}^{t = T}
  .
$$
This expression corresponds to the height difference
of the mass distribution
$
\bigl. m(Q) \bigr|_{Q = q(T)}^{Q = 0}
$
in the optimal schedule
(cf. Sec. \ref{sec:mass_distr}).
%
With a fixed $\alpha_0$,
increasing the simulation length $T$
also increases the optimal span $q(T)$ of the distribution
(cf. Sec. \ref{sec:optinitalpha}),
and thus reduces the height difference.
%
This effect is significant
if the distribution $m(Q)$ has a long tail,
or if some eigenvalues $\lambda_k$'s are near zero.
%
It is often suppressed in local sampling processes,
because the eigenmodes with large $\lambda_k$'s
are accompanied by large $\Gamma_k$'s,
which shortens the distribution tail.

>>>>

\title{Optimal updating factor in adaptive flat-distribution simulations}

@@

\subsubsection{\label{sec:nnscheme}
Triple-bin updating scheme}

@@

We tested the theory of computing errors
on the simple inverse-time schedule,
%
\begin{equation}
\alpha(t) = \frac{1}{\lambda \, (t + t_0) },
\label{eq:alpha_invtlambda}
\end{equation}
%
where $\lambda$ is a free parameter,
and $t_0 = 2/\alpha_0$.
%
The error of this schedule
can be computed analytically
(cf. Appendix \ref{sec:invt_schedule}),
and tested against simulation.

We performed FDS simulations
using the single-bin updating scheme
and the triple-bin updating scheme with $\mu_1 = 0.24$.
%
For each value of $\lambda$, we ran $1000$ independent simulations
of $10^8$ steps, and averaged the results.
%
As shown in Fig. \ref{fig:errinvt},
the errors from the simulations
agreed well those from the theory.

\begin{figure}[h]
\begin{center}
  \makebox[\linewidth][c]{
    \includegraphics[angle=0, width=1.0\linewidth]{fig/errinvt.pdf}
  }
  \caption{
    \label{fig:errinvt}
    Error $E$ versus $\lambda$
    in Eq. \eqref{eq:alpha_invtlambda}
    for the inverse-time schedule.
    %
    The points are results from averaging over $1000$ independent runs;
    the curves are theoretical predictions.
    %
    \note{The figure was produced by \texttt{doc/fig/errinvt.gp}.
      The data were produced by the script \texttt{data/invtcscan.py},
      and saved under \texttt{data/invt}
      (see the \texttt{Makefile} there).
    }%
  }
\end{center}
\end{figure}

For the single-bin updating scheme,
the optimal $\lambda$ occurred around $1.0$
for both global and local sampling processes.
%
The sampling process affected the error function
only by a multiplicative constant,
%The error functions of the two sampling processes
%differed only by a multiplicative constant,
in agreement with the discussion in
Appendix \ref{sec:invt_singlebin}.
%
For the triple-bin updating scheme,
the optimal $\lambda$ was less than $1.0$
for the global sampling process,
although it remained around $1.0$
for the local sampling process,
as discussed in Appendix \ref{sec:invt_nn}.
%
Thus, the inverse-time formula,
Eq. \eqref{eq:alpha_invt1},
is optimal for the single-bin updating scheme,
but not necessarily so for a general updating scheme.

@@

\subsection{\label{sec:results_cmpschemes}
Optimal updating schemes}

@@

%A practical problem of using homogeneous updating scheme
%is the following.
%
A homogeneous updating scheme
specified by an arbitrary kernel
is not necessarily stable,
i.e., some eigenvalues may be negative.
%
%
%\note{
  %For example,
  %the triple-bin updating scheme
  %is unstable if Eq. \eqref{eq:nn_stable} is violated.
  %
  Even
  the Gaussian updating scheme can be unstable
  if the updating kernel is truncated.
%}%
%
Here we show how to minimally modify
the updating kernel
to stabilize it.
%
\note{The relative code is \texttt{trimwindow()} in \texttt{invt.h}.
}

@@

For the single-bin scheme, we have
$\lambda_2 = \cdots = \lambda_n = 1$,
and
$\Gamma_k$'s affect the errors
%from Eqs. \eqref{eq:error_res_invt}
%and \eqref{eq:error_asym_invt}
only via the sum, $\sum_{k = 2}^n \Gamma_k$,
as a multiplicative constant.

@@

\subsection{\label{sec:invt_nn}
Triple-bin updating scheme}

For the triple-bin updating scheme,
%introduced in Sec. \ref{sec:nnscheme},
we can find an analytical expression
of the optimal value of $\lambda$
in the large-$n$ and asymptotic limit.
%
In this case,
if $2 \, \nu_k > 1$ for every $k$,
the asymptotic error is given by the integral
%$\sum_i \to \frac{2 \, n}{\pi} \int_0^{\pi/2} dp$, and
%
\begin{align}
  \Err_A
  =
  \frac{ 2 \, n  }
       { \pi \, T }
  \int_0^{\pi/2}
    \frac{ \Gamma(p) \, \left(1 - 4 \, \mu_1 \, \sin^2 p \right)^2    }
         {   \lambda \, \left(2 - 8 \, \mu_1 \, \sin^2 p - \lambda \right) }
  \, dp
  .
\notag
%\label{eq:error_nn_asym_int}
\end{align}

For perfect sampling,
$\Gamma(p) = 1$ from Eq. \eqref{eq:Gamma_perfect},
and
$$
\begin{aligned}
  \Err_A
  &=
  \frac{n}{4 \, T}
  \left(
    \frac{2 - 4 \, \mu_1 + \lambda}{ \lambda }
    +
    \frac{ \lambda }
    { \sqrt{ (2 - \lambda) (2 - 8 \, \mu_1 - \lambda) } }
  \right)
.
\end{aligned}
$$
%
Minimizing the function yields
%
\begin{equation}
  \lambda
  =
  \frac{ 1 - 4 \, \mu_1 }
       { 1 - 2 \, \mu_1 }
  ,
\notag
%\label{eq:lambda_nn_perfect}
\end{equation}
%
and the asymptotic error in the optimal case is given by
%
\begin{equation}
  \Err_A
  =
  \frac{n}{T}
  \left(
    1+ \frac{2 \, \mu_1^2}{1-4 \, \mu_1}
  \right)
  ,
\notag
%\label{eq:error_nn_perfect}
\end{equation}
%
%From Eq. \eqref{eq:error_nn}, it is clear that
which
is minimal at $\mu_1 = 0$, the single-bin case.
%increases with the magnitude of $\mu_1$
%no matter its sign,
%and the minimal value, obtained at $\mu_1 = 0$,
%corresponds to the single-bin scheme.
%

For the one-step sampling process,
we find from Eq. \eqref{eq:Gamma_onestep}
that $\Gamma(p) = \cot^2 p$
peaks around $p \approx 0$,
%
Then
$$
  \Err_A
  \propto
  \frac{   2 \, n }
       { \pi \, T }
  \frac{             1             }
       {  \lambda \, (2 - \lambda) }
  .
$$
%
The optimal $\lambda$ is, therefore,
%
$
%\begin{equation}
  \lambda \approx 1.
%\label{eq:lambda_nn_onestep}
%\end{equation}
$
%

\note{
However, the above result %Eq. \eqref{eq:lambda_nn_onestep}
is only good
for sufficiently short simulations,
since we have taken the limit of $n \to \infty$
for a fixed $T$.
%
For very long simulations,
the optimal $\lambda$
will ultimately drift to
the smallest eigenvalue,
$\min \lambda_k = 1 - 4 \, \mu_1$.
}
